---
title: "Week-2"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(tidyverse)
library(conflicted)
```

```{r}
install.packages("skimr")
library(skimr)

```





```{r}
#Reading the data set and printing first 5rows
data <- read.csv("dataset.csv")
head(data)
```

```{r}
# Display column names and types
str(data)
typeof(data)

#Verify
class(data)
```

The str() function shows the structure of the data (column names and types), while typeof() and class() provide additional type details of the entire dataset.

```{r}
conflicted::conflicts_prefer(dplyr::filter)

# Filter dataset where explicit is "True" and sample 10,000 rows
sample_data <- data |> filter(explicit == "True") |> sample_n(9000)
data <- sample_data
nrow(data)

# Display first few rows
head(data)
```


### Calculating summary statistics for popularity and danceability, including: Min/Max, Mean, Median, Quartiles, Standard deviations

```{r}
# 1. Numeric Summary for Columns - popularity and danceability

# Summary statistics using summary()
print("Summary statistics for 'popularity' and 'danceability':")
print(summary(data[c("popularity", "danceability")]))

# Quartiles for popularity
print("Quartiles for popularity:")
print(quantile(data$popularity, probs = seq(0, 1, 0.25)))

# Quartiles for danceability
print("Quartiles for danceability:")
print(quantile(data$danceability, probs = seq(0, 1, 0.25)))

# Standard deviation for both columns
print("Standard deviation of popularity:")
print(sd(data$popularity))

print("Standard deviation of danceability:")
print(sd(data$danceability))
```
### Using summarise() from dplyr to create a numeric summary for popularity and duration_ms

```{r}
# Numeric summary for two columns: popularity and duration_ms
summary(data$popularity)
summary(data$duration_ms)

data |>
  summarise(
    popularity_min = min(popularity, na.rm = TRUE),
    popularity_max = max(popularity, na.rm = TRUE),
    popularity_mean = mean(popularity, na.rm = TRUE),
    popularity_median = median(popularity, na.rm = TRUE),
    popularity_quantiles = quantile(popularity, probs = c(0.25, 0.5, 0.75), na.rm = TRUE),
    duration_min = min(duration_ms, na.rm = TRUE),
    duration_max = max(duration_ms, na.rm = TRUE),
    duration_mean = mean(duration_ms, na.rm = TRUE),
    duration_median = median(duration_ms, na.rm = TRUE),
    duration_quantiles = quantile(duration_ms, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
  )
```

```{r}
# 2. Categorical Summary (Unique Values & Counts)

# Count unique values in categorical columns
print("Number of unique values in 'track_genre' and 'explicit':")
print(sapply(data[c("track_genre", "explicit")], function(x) length(unique(x))))

# Frequency counts of each category in 'track_genre'
print("Frequency count of track genres:")
print(table(data$track_genre))

# Sorted frequency count of 'track_genre' in decreasing order
print("Sorted count of track genres (most common first):")
print(sort(table(data$track_genre), decreasing = TRUE))

# Frequency count of explicit (True/False)
print("Frequency count of explicit (True/False):")
print(table(data$explicit))
```

```{r}
# Function to show unique values and their counts for a categorical column
unique_values <- function(col) {
  # Create a frequency table
  freq_table <- table(col)
  # Convert the table into a data frame with 'Value' and 'Count'
  data.frame(
    Value = names(freq_table),
    Count.Freq = as.integer(freq_table)
  )
}

# Example usage for a categorical column like 'track_genre'
unique_values(data$track_genre)
# Example for multiple columns
unique_values(data$explicit)
unique_values(data$track_genre)
```

Combined Summary: This combines a numeric summary and a categorical summary (unique values) in one concise table.

```{r}
# 3. Combined Summary Using dplyr
summary_data <- data |>
  summarise(
    min_popularity = min(popularity),
    max_popularity = max(popularity),
    mean_popularity = mean(popularity),
    median_popularity = median(popularity),
    sd_popularity = sd(popularity),
    min_danceability = min(danceability),
    max_danceability = max(danceability),
    mean_danceability = mean(danceability),
    median_danceability = median(danceability),
    sd_danceability = sd(danceability),
    unique_genres = length(unique(track_genre)),
    unique_explicit = length(unique(explicit))
  )
print("Combined summary of numeric and categorical data:")
print(summary_data)
```
### 'skimr' provides an enhanced summary of the dataset, including the distribution and missing values

```{r}
print("Detailed summary using skimr:")
print(skim(data))
```

### Set of At Least 3 Novel Questions to Investigate:

1.  Does the popularity of songs vary significantly across genres?

2.  How does the duration of a song correlate with its popularity?

3.  Is there any connection between a song’s explicit content and its popularity or genre?

4.  Do explicit songs tend to have higher or lower popularity?

5.  How does the popularity of songs vary based on whether they are explicit or not?

6.  Do songs with longer durations have higher or lower popularity?

7.  Is there a trend in song popularity across different genres?

### Investigating Question using Aggregation 
```{r}
#Question 1: Does Popularity Vary by Genre?

# Aggregate popularity by genre
agg_genre <- data      |>
  group_by(track_genre) |>
  summarise(mean_popularity = mean(popularity, na.rm = TRUE))

print(agg_genre)
```


### Explanation: 
This aggregation finds the mean popularity for each genre. 

### Insight: 
The aggregated table will reveal if some genres are more popular on average. This insight can inform marketing strategies or targeted content creation. 

### Significance: 
If certain genres consistently have higher popularity scores, it may suggest genre preferences or trends that can be leveraged for recommendations or advertising. 

### Further Question: 
How do the distribution patterns compare for each genre?


### Data Visualization

```{r}
#Visualization 1: Distribution of Popularity (Histogram)

ggplot(data, aes(x = popularity)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Popularity", x = "Popularity", y = "Count")
```
### Explanation: 
The histogram shows the distribution of popularity scores across songs. 

### Insight: 
From the histogram, we can observe if popularity scores are normally distributed or if they are skewed towards higher or lower values.

### Significance: 
1. The histogram shows a right skew (more songs with lower popularity), it suggests that only a few songs reach high popularity. This could be useful for modeling popularity prediction. 

### Further Question: 
Are there significant peaks in the popularity distribution that could indicate genre-specific trends?

```{r}
#Visualization:2 Scatterplot using Energy and Loundness

ggplot(sample_data, aes(x = energy, y = loudness)) +
  geom_point(alpha = 0.5, color = "green") +
  labs(title = "Energy vs Loudness", x = "Energy", y = "Loudness (dB)") +
  theme_minimal()
```

### Explanation:
This scatter plot explores the relationship between a song’s energy and its loudness.

### Insights:
1. A strong positive correlation suggests that louder tracks tend to have higher energy.
2. If points are scattered without a clear trend, it implies that loudness and energy are independent.
3. Outliers in the high-energy, low-loudness region could indicate high-energy but soft acoustic tracks.

### Significance:
Understanding this relationship helps identify how music production techniques influence perception—high-energy tracks with low loudness could be a unique characteristic of certain genres.

### Further Questions:
1. Are there specific genres that break the typical loudness-energy relationship?
2. Does explicit content correlate with louder and more energetic songs?

```{r}
# Visualisation: 3 Density plot of Acousticness

ggplot(sample_data, aes(x = acousticness)) +
  geom_density(fill = "purple", alpha = 0.5) +
  labs(title = "Density Plot of Acousticness", x = "Acousticness", y = "Density") +
  theme_minimal()
```

### Explanation:
This density plot shows the probability distribution of acousticness values in the dataset.

### Insights:
A peak near 0 indicates many tracks are highly acoustic.
A bimodal or multimodal distribution suggests there are distinct clusters of acoustic and non-acoustic tracks.
A left-skewed distribution means most tracks are electronic or heavily processed.

### Significance:
Acousticness is an important factor in determining whether a song is more organic or digitally produced.

### Further Questions:
Do explicit tracks tend to have lower acousticness compared to clean tracks?
How does acousticness relate to energy and tempo?
